---
title: "Exploratory Data Analysis of Tooth Growth in Gerbils"
author: "AKN"
date: "Sunday, September 27, 2015"
output:
  pdf_document:
    fig_caption: yes
    fig_height: 3
    keep_tex: yes
    latex_engine: xelatex
  html_document: default
---
#Overview
In this report, we use `ToothGrowth`, a dataset in R, which contains data from an experiment measuring the length of guinea pig teeth, or _odontoblasts_, at three dose levels of Vitamin C, either through Orange Juice or with just ascorbic acid. We will perform exploratory data analysis, followed by hypothesis testing and calculate confidence intervals. Finally, we will show that the tooth growth does increase with increasing dose of Vitamin C.

```{r setup, cache=TRUE, echo=FALSE, results='hide', warning=FALSE, message=FALSE}
library(ggplot2)
library(reshape2)
library(knitr)
library(dplyr)
library(gridExtra)
```

#The Data and Exploratory Analyses
First, we load the ToothGrowth data. The data contains 60 observations in 3 dose levels of 2 supplement types. First, we examine how the response variable, tooth length, varies by dose level and supplement type. We can do this with a simple boxplot.  

```{r loading, dependson=1, cache=TRUE}
data1 <- ToothGrowth
data1$dose <- as.factor(data1$dose)
```

```{r plots, dependson=c(1,2), cache=TRUE, echo=FALSE,fig.width=10, fig.width=6, fig.cap="Data Summary"}
p <- ggplot(data=data1, aes(y=len,x=dose, fill=supp)) + geom_boxplot() 
p <- p + labs(x = "Dose level(mg)", y = "Tooth Length")
p <- p + scale_fill_discrete(name=" ",breaks=c("OJ", "VC"),
                                   labels=c("Orange Juice", "Ascorbic Acid")) 
p <- p +theme(legend.position="bottom")
grid.arrange(p, tableGrob(summary(data1),cols = c("Length", "Supplement", "Dose")), ncol=2)
```  
  
  Visually, it looks like Tooth Length is longer at higher dose levels, regardless of the supplement type. However, there may be some effect of supplement type as well. The experimental design of this study essentially has two levels. So, we can group the data in few ways, only by supp, only by dose, or by both dose and supplement. We can take a look at the subgroup means and variances before we get to hypothesis testing.  
  
```{r table, dependson=c(1,2), cache=TRUE, echo=FALSE, fig.width=10, fig.cap="Tables of Group Means/Variances"}
datasupp <- group_by(data1, supp) %>% summarise(mean=mean(len), var=var(len))
datadose <- group_by(data1, dose) %>% summarise(mean=mean(len), var=var(len))
data2 <- group_by(data1, dose, supp) %>% summarise(mean=mean(len),var = var(len))
grid.arrange(tableGrob(datadose), tableGrob(datasupp), tableGrob(data2), ncol=3)
```  
In this report, we will use t-test and the calculated confidence intervals to determine if there is a difference between the mean tooth growth at different dose and type levels. 
    
#Confidence Intervals and Hypothesis Testing  

Let's look at the supplement type first. First we compare the mean and variance at each supplement type. Referring to our tables of means and variances above, we see that the variances are `43.6` and `68.32`. To be safe, we set `var.equal = FALSE`, in our t-test. We know from the description of the dataset that the groups are not paired.  

```{r ttest, dependson=c(1,2), cache=TRUE, echo=TRUE}
t.test(len ~ supp,data=data1, var.equal = FALSE, paired = FALSE)
```
Although the p-value is `0.06`, the confidence interval contains `0`. Because the 95% CI includes 0, we cannot reject the null hypothesis that the difference in means is 0.
Next, we will test the group differences by dose level, using `t.test` as before. Again, to be safe, we assume the group variances are not equal.  

```{r CI, dependson=c(1,2),cache=TRUE, fig.width=8}
dlong <- melt(data1, id.vars = c("dose"), measure.vars="len")
g1 <- dlong[dlong$dose == "0.5",]$value
g2 <- dlong[dlong$dose == "1",]$value
g3 <- dlong[dlong$dose == "2",]$value
t1 <- t.test(g2, g1,var.equal = FALSE, paired = FALSE)
t2 <- t.test(g3, g1,var.equal = FALSE, paired = FALSE)
t3 <- t.test(g3, g2,var.equal = FALSE, paired = FALSE)
conf_int <- rbind(t1$conf, t2$conf, t3$conf)
p_vals <- rbind(t1$p.value, t2$p.value, t3$p.value)
t_stat <- rbind(t1$statistic, t2$statistic,t3$statistic)
labels <-c("1mg vs. 0.5mg", "2mg vs. 0.5mg", "2mg vs 1mg")
results1 <- data.frame(labels, conf_int, p_vals, t_stat)
colnames(results1) <- c("Comparison", "95% CI", "95% CI", "p", "t")
kable(results1, digits = 3)
```
  
  We have now done 4 successive t-tests on the data, comparing various combinations of groups. But what about power? We have estimates of the 'true' mean difference and its variation from our data.
  
```{r power, dependson=c(1,2), cache=TRUE}
meandiff <- c(mean(g3)-mean(g2), mean(g3)-mean(g1), mean(g2-g1))
vars1 <- c(var(g1), var(g2), var(g3))
sd1 <- sqrt(vars1)
rbind(meandiff, sd1)
power.t.test(n=10, type = "two.sample", alternative = "two.sided", sig.level = 0.01, sd=4.5, delta=6.4)$power
```  
  At an of 10, and $\alpha$ of 0.01 we have 62% power to correctly reject a null hypothesis assuming a true mean difference of ~6.4 (smallest diff observed). 
  
#Conclusions
  1. When comparing the means at each dose level (0.5, 1 and 2 mg), all 3 combinations of comaprisons produce confidence intervals that do not contain 0, t > 1, and p < 0.001.
  2. We infer from this data that while the *type* of Vitamin C supplement *does not* significantly affect tooth growth, higher *doses* of Vitamin C *do increase* tooth growth.
  3. We may be underpowered, as we are at <75% power to correctly reject a null hypothesis when the true mean difference is equal to the smallest difference we observed.
    
#Assumptions  
  1. We assumed that the variances across the group (either by dose or supplement) was not equal. 
  2. We assumed that the 60 gerbils used for this data were randomy assigned to each of the dose/supplement groups.
  3. We assumed that the 60 gerbils are representative of the entire population of gerbils. 
  4. Finally, we assumed that the response variable is not itself affected by other variables that we have not included in our statistical inference. 
